---
title: "Generalized Additive Models: High severity in relation to climate environment"
author: "S. Haire"
date: "7/22/2020"
output: html_document
---
The purpose of the script is to run Generalized Additive Models for the high severity threshold data and plot the results.

Just the first principal component is analyzed here. 

### Notes: 
For development of input data, see: severity.thresholds.Rmd

The metric of interest for this analysis is pland=Percent of the landscape (composition of high severity)

pc1 = first principal component

Code for GAMs, CIs and plots adapted from:
Simpson, Gavin L. 2018. Modelling Palaeoecological Time Series Using Generalised Additive Models. Frontiers in Ecology and Evolution 6: 149. https://doi.org/10.3389/fevo.2018.00149.

```{r setup, include=FALSE}

## might not need all these!
library(ggplot2)
library(ggpubr)
library(cowplot)
library(plyr)
library(dplyr)
library(mgcv)
library(scam)
library(tidyr)
library(tidyverse)
#install.packages("devtools")
#devtools::install_github("gavinsimpson/gratia")
library(gratia)

```

### Read in the data and prepare for analysis


```{r dataprep, include=TRUE}
datap<-"./data/tabular/" # where the input datafile is located

## pland data for all fires
pland.hi<-read.csv(paste(datap, "class.metrics.high.csv",sep="")) 

## rescale pc values to 0,1 
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
pland.hi$pc1<-range01(pland.hi$pc1)  
pland.hi$pc2<-range01(pland.hi$pc2)

```

### Analysis of Mexico data

```{r mexico, echo=FALSE}
mx.hi<-pland.hi[pland.hi$Country=="MEX",]
mod.mx <- gam(pland ~ s(pc1, k = 12), data = mx.hi, method = "REML")
## take a look at gam results
summary(mod.mx)
plot(mod.mx, main="Mexico")

### model evaluation
## 
N <- 1000 # number of points at which to evaluate the smooth
## create new data to predict at; 1000 evenly-spaced values over `pc1`
new.pc1.mx<- with(mx.hi, data.frame(pc1 = seq(min(pc1), max(pc1), length.out = N)))
## Predict from the fitted model; note we predict from the $gam part
new.pc1.mx <- cbind(new.pc1.mx, data.frame(predict(mod.mx, new.pc1.mx, se.fit = TRUE)))
new.pc1.mx$Method<-rep("GAM (GCV)", N) # not using this...
## Create the confidence interval
crit.t <- qt(0.95, df = df.residual(mod.mx))
new.pc1.mx <- transform(new.pc1.mx,
                        upper = fit + (crit.t * se.fit),
                        lower = fit - (crit.t * se.fit))

## Plot estimated trend w ribbons
xlim=c(min(pland.hi$pc1), max(pland.hi$pc1))
ylim=c(min(pland.hi$pland), max(pland.hi$pland))
high_fitted.mx.pc1<-ggscatter(mx.hi, "pc1", "pland", color="threshold.value",   xlab="PC1",ylab="High severity (%)",main="", xlim=xlim, ylim=ylim) + theme_gray() + gradient_color("black") + geom_ribbon(data=new.pc1.mx, mapping=aes(x=pc1, ymin = lower, ymax = upper), alpha = 0.1, inherit.aes = FALSE) +  geom_line(data=new.pc1.mx, mapping=aes(x=pc1, y=fit)) + theme(legend.position ="right") + theme(text = element_text(size = 15))
  
```

### Analysis of US data

```{r usa, echo=FALSE}

## US
us.hi<-pland.hi[pland.hi$Country=="US",]
mod.us <- gam(pland ~ s(pc1, k = 12), data = us.hi, method = "REML")
## check gam results
summary(mod.us)
plot(mod.us, main="US")
## model evaluation
N <- 1000 # number of points at which to evaluate the smooth
## create new data to predict at; 1000 evenly-spaced values over `pc1`
new.pc1.us<- with(us.hi, data.frame(pc1 = seq(min(pc1), max(pc1), length.out = N)))
new.pc1.us <- cbind(new.pc1.us, data.frame(predict(mod.us, new.pc1.us, se.fit = TRUE)))
new.pc1.us$Method<-rep("GAM (GCV)", N) # not using this
## Create the confidence interval
crit.t <- qt(0.95, df = df.residual(mod.us))
new.pc1.us <- transform(new.pc1.us,
                           upper = fit + (crit.t * se.fit),
                           lower = fit - (crit.t * se.fit))

## Plot estimated trend w ribbons
xlim=c(min(pland.hi$pc1), max(pland.hi$pc1))
ylim=c(min(pland.hi$pland), max(pland.hi$pland))
high_fitted.us.pc1<-ggscatter(us.hi, "pc1", "pland", color="threshold.value", xlab="PC 1",ylab="High severity (%)", main="", xlim=xlim, ylim=ylim) + theme_gray() + gradient_color("black") + geom_ribbon(data=new.pc1.us, mapping=aes(x=pc1, ymin = lower, ymax = upper), alpha = 0.1, inherit.aes = FALSE) +  geom_line(data=new.pc1.us, mapping=aes(x=pc1, y=fit)) +  theme(legend.position ="right")+ theme(text = element_text(size = 15))

```

```{r plots, echo=FALSE}
plot_grid(high_fitted.us.pc1, high_fitted.mx.pc1,ncol =1,align ="hv",axis ="lr", labels=c('US', 'Mexico'))
```

